Timing and Positioning: Insights from Steak-Related Product Pricing at Walmart and Loblaws

### 2024-12-03 paper/paper.qmd
Update paper.qmd

---
title: "Timing and Positioning: Insights from Steak-Related Product Pricing at Walmart and Loblaws"
subtitle: "My subtitle if needed"
author: 
  - Wei Wang
thanks: "Code and data are available at: [https://github.com/zcyjn233/Explore-steak-related-products)."
date: today
date-format: long
abstract: "This study explores pricing strategies in the retail grocery sector, with a focus on steak-related products from Walmart and Loblaws. Utilizing a comprehensive dataset of historical and current prices, alongside detailed product descriptions and timing of price adjustments, the research identifies significant patterns influencing consumer behavior. Advanced statistical models and visual analyses provide insight into pricing dynamics and highlight differences in strategy between the two chains. The findings reveal actionable trends, emphasizing the role of timing and product positioning in shaping customer purchasing decisions.."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The results of the model analysis reveal several key insights into the pricing dynamics of steak-related products. Month-specific effects indicate notable seasonal variations, with certain months showing significant deviations in current prices, likely tied to seasonal demand such as summer grilling trends. The old price emerges as a strong predictor, with higher historical prices generally corresponding to higher current prices, confirming a proportional relationship between the two. Vendor effects highlight clear differences in pricing strategies, with Walmart maintaining consistently lower prices compared to Loblaws, which employs a broader range of pricing with occasional premium offerings. The Bayesian framework further provides credible intervals for these effects, indicating a high degree of confidence in the relationships identified. Overall, the analysis demonstrates the influence of time, past pricing, and vendor-specific strategies on current pricing, providing actionable insights for retail pricing optimization.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

@@

The analysis highlights the distinct pricing strategies employed by Walmart and Loblaws, shedding light on how these retailers align their approaches with their brand identities and target audiences. Walmart's consistent low-pricing strategy appeals to budget-conscious consumers, maintaining steady demand throughout the year. Conversely, Loblaws employs a more dynamic pricing strategy, utilizing seasonal promotions and premium offerings to attract a broader demographic, including shoppers seeking high-quality or specialty products.

A key discovery is the proportional relationship between old and current prices, suggesting that historical pricing not only reflects past demand but also anchors consumer expectations and maintains market stability. While Walmart’s approach provides predictability, Loblaws’ flexibility in pricing allows it to capitalize on periods of heightened demand, such as during the summer grilling season. This contrast underscores the importance of tailoring pricing strategies to align with both market conditions and brand positioning. Retailers can use these insights to refine their pricing approaches, balancing stability with responsiveness to seasonal trends and consumer preferences.

@@

Seasonality emerged as a major driver of pricing adjustments, with notable fluctuations in months such as June and July, aligning with peak grilling periods. The data reveal that pricing strategies during these months capitalize on increased consumer demand for steak-related products, demonstrating the importance of timing in retail pricing decisions. For Walmart, steady pricing minimizes seasonal volatility, while Loblaws uses targeted promotions to drive seasonal sales spikes. This finding emphasizes the importance of aligning pricing strategies with consumer behavior, allowing retailers to maximize revenue during high-demand periods.

A new insight from the data is the interplay between promotional pricing and consumer psychology. Significant markdowns, such as steep discounts on premium products like ribeye steak, drive temporary demand spikes, indicating that perceived value plays a critical role in consumer purchasing decisions. This suggests that well-timed promotions not only boost short-term sales but also create opportunities for customer engagement. Retailers should explore the thresholds at which discounts maximize consumer response without eroding long-term profitability.
## Implications, Limitations, and Future Directions

This study provides actionable insights into retail pricing, but it also highlights areas for further exploration. The model effectively integrates prior knowledge with observed data, offering robust predictions and uncovering key trends. However, limitations include the inability to fully capture outlier behaviors, such as aggressive discounting on high-value products, and the reliance on historical prices, which may oversimplify the influence of real-time factors like advertising or supply chain disruptions. Addressing these limitations through non-linear models or incorporating external variables, such as competitor pricing or macroeconomic trends, could enhance the model’s precision.

Future research should focus on refining models to account for regional and demographic differences, offering more granular insights into consumer behavior. Additionally, exploring the long-term impacts of pricing strategies on customer loyalty would provide valuable guidance for balancing short-term revenue boosts with sustainable growth.

Overall, this study underscores the critical role of pricing in shaping competitive dynamics in the retail grocery sector. By leveraging insights into vendor strategies, seasonal trends, and consumer behavior, retailers can optimize their pricing decisions to drive sales, build loyalty, and maintain a competitive edge in an ever-evolving market.
## Weaknesses and next steps

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows The posterior predictive check assesses how well the model represents the observed data by comparing the predicted data distribution to the actual data. The analysis shows that the model captures the main features of the observed price distribution, including its central tendencies and variability. The close alignment between simulated and actual data distributions indicates that the model is robust and reliable in representing the pricing trends of steak-related products. However, minor deviations at the extremes suggest potential opportunities for model refinement, particularly in handling outlier cases. Overall, the predictive check confirms that the model is well-suited to capture the key pricing dynamics in the dataset.

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows The comparison between posterior and prior distributions demonstrates how the data influences the model’s parameter estimates. Key parameters such as vendor-specific effects and historical prices show significant shifts from their prior distributions, indicating that the data strongly informs these estimates. This aligns with findings that vendor-specific pricing strategies and historical prices are critical factors in determining current prices. Conversely, parameters such as the intercept show minimal deviation, reflecting limited evidence from the data to adjust initial assumptions. This balance underscores the model's ability to incorporate prior knowledge while letting the data drive meaningful updates.

Together, these analyses validate the model's capability to capture real-world pricing trends, providing actionable insights into how vendors, historical prices, and seasonal factors influence current pricing strategies. These results reinforce the model's value for guiding competitive pricing decisions in the retail grocery market.

@@

@fig-stanareyouokay-1 is a trace plot.The trace plot, shown in Figure 1, visualizes the Markov Chain Monte Carlo (MCMC) sampling process for each parameter across four chains. The chains mix well and converge toward a stationary distribution, indicated by the consistent overlap of the sampled values over time. This pattern suggests that the sampling process has stabilized, providing reliable parameter estimates. Additionally, the absence of irregularities or trends in the trace lines confirms that the model has reached convergence.

@fig-stanareyouokay-2 is a Rhat plot. It shown in Figure 2, assesses the convergence of the MCMC algorithm by comparing within-chain and between-chain variability for each parameter. All Rhat values are very close to 1, indicating that the chains have converged, and the posterior estimates are stable. There are no signs of divergence or poor mixing across chains, confirming the validity of the model's MCMC sampling process.

Together, these diagnostics demonstrate that the MCMC algorithm has successfully converged, ensuring that the parameter estimates are robust and suitable for inference.
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

>>>>

---
title: "Timing and Positioning: Insights from Steak-Related Product Pricing at Walmart and Loblaws"
subtitle: "Insights into Pricing Strategies: Unveiling Research Results and the Key Factors Driving Steak-Related Product Pricing"
author: 
  - Wei Wang
thanks: "Code and data are available at: [https://github.com/zcyjn233/Explore-steak-related-products)."
date: today
date-format: long
abstract: "This study explores pricing strategies in the retail grocery sector, with a focus on steak-related products from Walmart and Loblaws. Utilizing a comprehensive dataset of historical and current prices, alongside detailed product descriptions and timing of price adjustments, the research identifies significant patterns influencing consumer behavior. Advanced statistical models and visual analyses provide insight into pricing dynamics and highlight differences in strategy between the two chains. The findings reveal actionable trends, emphasizing the role of timing and product positioning in shaping customer purchasing decisions.."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The results of the Bayesian model analysis provide a comprehensive understanding of the pricing dynamics of steak-related products, shedding light on how different factors interact to influence current pricing. One of the most significant findings is the presence of strong month-specific effects, which highlight the importance of seasonality in pricing strategies. Certain months, such as those during the summer grilling season, show notable deviations in current prices, reflecting heightened consumer demand for steak-related products during this period. This seasonal trend aligns with the broader market behavior, where retailers adjust prices to capitalize on increased consumer interest in specific product categories during peak times. Such findings emphasize the importance of incorporating time-sensitive factors into retail pricing models to better predict consumer behavior and optimize revenue.

Historical pricing, represented by the old price variable, emerges as another strong predictor of current prices. The analysis demonstrates a proportional relationship between old and current prices, where higher historical prices generally correspond to higher current prices. This finding suggests that past pricing not only reflects demand patterns but also serves as a baseline or anchor for setting future prices. Retailers appear to use historical prices as a reference point to maintain consistency, manage consumer expectations, and preserve brand positioning. This relationship underscores the significance of historical pricing data in understanding how retailers balance competitiveness with profitability.

The model also highlights vendor-specific effects, revealing clear differences in the pricing strategies employed by Walmart and Loblaws. Walmart's approach, characterized by consistently lower prices and a narrower range of variability, aligns with its brand image as a value-oriented retailer catering to price-sensitive consumers. On the other hand, Loblaws exhibits a broader pricing range, incorporating premium pricing strategies and occasional discounts to attract a more diverse demographic, including those seeking higher-quality or specialty products. This contrast in strategies reflects the competitive dynamics within the grocery sector, where retailers adopt distinct approaches to differentiate themselves and capture their target markets.

The Bayesian framework used in the analysis enhances the robustness of these findings by providing credible intervals for all estimated effects. These intervals quantify the uncertainty associated with the relationships identified, offering a nuanced understanding of the strength and variability of each predictor's influence. For example, while the month-specific effects show significant deviations in certain periods, the credible intervals confirm the reliability of these findings across different scenarios. Similarly, the strong association between old and current prices is supported by tight credible intervals, indicating a high degree of confidence in this relationship. The vendor-specific effects also demonstrate consistent patterns, further validating the distinct pricing approaches of Walmart and Loblaws.

Overall, the analysis illustrates how time, historical pricing, and vendor strategies converge to shape current pricing. The findings provide actionable insights for retail pricing optimization, suggesting that retailers can leverage historical data and seasonal trends to better align their pricing strategies with consumer behavior. Additionally, understanding the differences in vendor-specific approaches offers valuable lessons for market positioning, highlighting the need to tailor pricing strategies to a retailer's target audience and competitive environment. By combining these insights with the flexibility and precision of Bayesian modeling, the study offers a robust framework for exploring and optimizing pricing dynamics in the retail sector.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory model of current prices based on month, old price, and vendor effects."
#| warning: false

@@

The analysis reveals valuable insights into the distinct pricing strategies employed by Walmart and Loblaws, offering a clearer understanding of how these retailers align their approaches with their respective brand identities and target audiences. Walmart has adopted a consistent low-pricing strategy that prioritizes affordability and accessibility, appealing to budget-conscious consumers. This approach ensures steady demand throughout the year and reinforces Walmart’s reputation as a cost-effective retailer. By keeping price variability low, Walmart creates a predictable shopping experience that fosters customer loyalty and positions itself as a reliable choice for price-sensitive shoppers.

In contrast, Loblaws employs a more dynamic pricing strategy that incorporates seasonal promotions, targeted discounts, and premium offerings. This flexibility allows Loblaws to attract a diverse demographic, including consumers who seek higher-quality products or are enticed by promotional deals. Seasonal price adjustments, particularly during high-demand periods like the summer grilling season, enable Loblaws to capitalize on increased interest in steak-related products, driving sales and expanding its customer base. The dynamic nature of Loblaws’ pricing strategy also positions the retailer as an adaptable and competitive player in the market, capable of responding to changing consumer demands and market conditions.

A key discovery from the analysis is the proportional relationship between historical prices (old prices) and current prices. This relationship suggests that historical pricing serves as an anchor for setting future prices, helping retailers manage consumer expectations and maintain market stability. Walmart leverages this relationship to reinforce its low-price reputation, ensuring that customers perceive consistent value in their purchases. On the other hand, Loblaws uses historical prices to balance promotional markdowns with premium pricing, preserving the perceived value of its offerings even during discount periods.

These findings underscore the importance of tailoring pricing strategies to align with market conditions, brand positioning, and consumer preferences. Retailers can draw on these insights to refine their pricing approaches, balancing stability with responsiveness to seasonal trends and competitive pressures. By understanding how historical prices, vendor strategies, and seasonal trends interact, retailers can optimize their pricing decisions to drive sales, enhance customer satisfaction, and maintain a competitive edge in the market. Additionally, these findings highlight the value of data-driven pricing strategies in navigating the complexities of consumer behavior and market dynamics.

@@

Seasonality emerged as one of the most significant drivers of pricing adjustments in the analysis, with notable price fluctuations during months such as June and July. These months align with peak grilling periods, when consumer demand for steak-related products tends to be at its highest. Retailers strategically adjust their pricing during these times to capitalize on increased demand, demonstrating the importance of timing in retail pricing decisions. For instance, Loblaws uses targeted promotions during these months to attract consumers, driving seasonal sales spikes and maximizing revenue opportunities. In contrast, Walmart’s consistent low-pricing strategy minimizes seasonal volatility, maintaining a steady appeal to budget-conscious shoppers throughout the year.

An important insight from the analysis is the interplay between promotional pricing and consumer psychology. Significant markdowns, such as steep discounts on premium products like ribeye steak, create temporary spikes in demand, suggesting that perceived value plays a critical role in consumer purchasing decisions. This behavior highlights the effectiveness of well-timed promotions in boosting short-term sales and engaging customers. Retailers can use this understanding to design promotions that not only attract shoppers but also enhance the overall shopping experience, fostering long-term customer loyalty. However, this strategy requires careful calibration to ensure that discounts do not erode profitability or diminish the perceived value of premium products.

The findings also emphasize the importance of aligning pricing strategies with consumer behavior patterns to maximize revenue during high-demand periods. For example, leveraging seasonal trends to introduce promotional discounts or premium bundles can increase consumer spending while enhancing brand differentiation. Retailers should explore the thresholds at which discounts generate the maximum response from customers without compromising long-term revenue. Additionally, understanding the psychological drivers of consumer behavior can help retailers fine-tune their pricing strategies to appeal to different market segments, from bargain hunters to premium shoppers.

## Implications, Limitations, and Future Directions

This study provides actionable insights into retail pricing dynamics while also highlighting areas for improvement and future exploration. The Bayesian model used in the analysis effectively integrates prior knowledge with observed data, producing robust predictions and uncovering key trends. It highlights the importance of vendor-specific strategies, historical pricing, and seasonal trends in shaping current prices, offering a framework for optimizing pricing decisions in a competitive market. However, the analysis also identifies several limitations that warrant further attention.

One limitation is the model’s inability to fully capture outlier behaviors, such as aggressive discounting on high-value products. These behaviors, while rare, can significantly influence consumer perceptions and market dynamics. Addressing this limitation may require the use of non-linear models or interaction terms to better capture the complexities of pricing strategies. Another limitation is the reliance on historical prices, which may oversimplify the influence of real-time factors such as advertising, competitor pricing, or supply chain disruptions. Incorporating these factors into the model could enhance its precision and relevance, providing a more comprehensive understanding of retail pricing dynamics.

Future research should focus on refining the model to account for regional and demographic differences, offering more granular insights into consumer behavior. For example, analyzing how pricing strategies vary across different geographic regions or consumer segments could uncover additional patterns that inform targeted marketing and pricing decisions. Additionally, exploring the long-term impacts of pricing strategies on customer loyalty and brand equity would provide valuable guidance for balancing short-term revenue boosts with sustainable growth. This could include investigating the effects of promotional campaigns on repeat purchases or the role of premium pricing in enhancing brand perception.

Overall, this study underscores the critical role of pricing in shaping competitive dynamics in the retail grocery sector. By leveraging insights into vendor strategies, seasonal trends, and consumer behavior, retailers can optimize their pricing decisions to drive sales, build customer loyalty, and maintain a competitive edge. The findings also highlight the importance of adopting flexible, data-driven approaches to pricing, enabling retailers to adapt to changing market conditions and consumer preferences. As the retail landscape continues to evolve, these insights provide a valuable foundation for developing innovative and effective pricing strategies.
## Weaknesses and next steps

@@

### Posterior Predictive Check (@fig-ppcheckandposteriorvsprior-1)
The posterior predictive check evaluates how well the model represents the observed data by comparing the distribution of predicted data (posterior simulations) to the actual data. This assessment is crucial for determining whether the model captures the core characteristics of the observed data, including its central tendencies, variability, and shape. The analysis demonstrates that the model successfully replicates the main features of the observed price distribution for steak-related products, indicating that it is robust and reliable in capturing the overall pricing trends.

The close alignment between the simulated predictions and the actual data distributions reflects the model's ability to generalize well to the underlying patterns in the dataset. However, the analysis also reveals minor deviations at the extremes of the distribution, suggesting areas where the model could benefit from refinement. Specifically, these discrepancies may point to outlier cases or unusual pricing behaviors that are not fully captured by the current model structure. For example, aggressive discounting or premium pricing on specialty products might contribute to these deviations, highlighting the potential for including additional factors or non-linear effects in the model.

Overall, the posterior predictive check confirms that the model is well-suited to capture the key pricing dynamics present in the dataset. The results validate the appropriateness of the Bayesian approach for understanding pricing strategies while also providing a roadmap for further enhancements to address edge cases. By successfully balancing the incorporation of prior knowledge with observed data, the model proves itself as a reliable tool for investigating the pricing mechanisms of steak-related products across Walmart and Loblaws.

### Prior-Posterior Comparison (@fig-ppcheckandposteriorvsprior-2)
The comparison between posterior and prior distributions demonstrates how the observed data informs the model’s parameter estimates. Key parameters, such as vendor-specific effects and historical prices, show substantial shifts from their prior distributions to the posterior distributions. This indicates that the data strongly influences these parameters, confirming their significance in predicting current prices. For example, the posterior distribution for vendor-specific effects highlights the distinct pricing strategies of Walmart and Loblaws, aligning with findings that these strategies play a critical role in shaping consumer behavior and pricing trends.

Conversely, parameters like the intercept show minimal deviation between their prior and posterior distributions. This reflects limited evidence from the data to adjust initial assumptions about baseline prices. Such consistency suggests that the prior assumptions for these parameters were well-calibrated, reinforcing the importance of informed priors in Bayesian modeling. Additionally, the prior-posterior comparison demonstrates how the Bayesian framework allows the model to balance prior knowledge with data-driven updates, ensuring that the final estimates are both robust and interpretable.

This analysis underscores the flexibility of the Bayesian approach in refining parameter estimates based on observed data while maintaining a foundation of prior beliefs. By integrating these two sources of information, the model delivers meaningful insights into the factors driving current pricing strategies, including vendor behaviors, historical prices, and seasonal trends.

### Combined Insights

Together, the posterior predictive check and prior-posterior comparison validate the model’s ability to capture real-world pricing trends. The strong alignment between simulated and observed data, coupled with data-driven adjustments to key parameters, highlights the robustness of the model in representing the complexities of steak-related product pricing. These findings provide actionable insights into how vendors, historical prices, and seasonal factors influence pricing strategies, reinforcing the value of the model as a decision-making tool for retailers.

The results also emphasize the importance of using advanced modeling techniques like Bayesian regression to uncover nuanced patterns in retail data. By allowing for the integration of prior knowledge and observed data, the Bayesian framework provides a flexible and powerful approach to understanding and predicting pricing dynamics. These insights are critical for guiding competitive pricing decisions in the retail grocery market, enabling retailers to optimize their strategies and better align with consumer behavior.

@@

### Trace Plot (@fig-stanareyouokay-1)
The trace plot, shown in Figure 1, provides a visual representation of the Markov Chain Monte Carlo (MCMC) sampling process for each parameter across four chains. It serves as a critical diagnostic tool to assess whether the chains are mixing well and converging toward a stationary distribution. In this analysis, the trace lines for all parameters exhibit consistent overlap over time, with no visible trends, drifts, or irregularities. This indicates that the chains are exploring the parameter space effectively and have stabilized.

The stability of the trace lines suggests that the MCMC algorithm has reached convergence, meaning that subsequent samples are being drawn from the posterior distribution rather than being influenced by the starting points of the chains. The absence of abrupt jumps or prolonged trends in the trace lines further confirms that the sampling process is reliable, providing robust parameter estimates. This level of consistency is essential for ensuring the credibility of the model's outputs, as it eliminates concerns about biased or incomplete exploration of the posterior space.

Additionally, the trace plot demonstrates the reproducibility of the sampling process, as all four chains show similar behavior across iterations. This reinforces confidence in the robustness of the Bayesian model and its ability to provide accurate insights into the pricing dynamics of steak-related products. By visualizing the mixing and stability of the chains, the trace plot effectively validates the reliability of the MCMC process and supports the interpretation of the model’s posterior estimates.

### Rhat Plot (@fig-stanareyouokay-2)
The Rhat plot, shown in Figure 2, quantitatively assesses the convergence of the MCMC algorithm by comparing within-chain and between-chain variability for each parameter. The Rhat diagnostic, also known as the Gelman-Rubin statistic, measures whether the chains have mixed well and converged to the same posterior distribution. In this analysis, all Rhat values are very close to 1, indicating that the chains have converged successfully, and the posterior estimates are stable.

A value of Rhat close to 1 suggests that the variance within each chain is consistent with the variance between chains, confirming that all chains are sampling from the same posterior distribution. This eliminates concerns about poor mixing or divergence, ensuring that the MCMC algorithm has fully explored the parameter space. Importantly, the absence of any parameters with Rhat values above the commonly accepted threshold of 1.05 further validates the reliability of the sampling process.

The Rhat plot complements the trace plot by providing a quantitative metric for assessing convergence. While the trace plot offers a visual confirmation of stable mixing, the Rhat plot provides numerical evidence that all chains have successfully converged. Together, these diagnostics demonstrate that the MCMC process is robust and that the resulting posterior estimates are suitable for inference.

### Combined Insights
Together, the trace and Rhat plots validate the success of the MCMC algorithm in producing reliable parameter estimates. The trace plot confirms visually that the chains mix well and converge to a stationary distribution, while the Rhat plot quantitatively reinforces this conclusion by demonstrating consistent variability within and between chains. The alignment of these diagnostics eliminates any concerns about biased or incomplete exploration of the posterior distribution.

These diagnostics are critical for ensuring the robustness of the Bayesian model, as convergence is a prerequisite for drawing meaningful inferences from the posterior estimates. By confirming the stability and reliability of the MCMC process, the trace and Rhat plots enhance confidence in the model’s ability to accurately capture the pricing dynamics of steak-related products. This ensures that the insights derived from the model are not only statistically sound but also actionable for guiding pricing strategies in the retail grocery sector.
```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2
